---
title: "Fixing Parameters Before Calibration"
output: distill::distill_article
bibliography: vignette_phenoflex.bib
csl: elsevier-harvard-modifed.csl
---

Fixing model parameters to default values—so that they are excluded from calibration—is relatively straightforward in PhenoFlex, thanks to the modular design of the `chillR::phenologyFitter()` function. The key is to modify the `modelfn` argument, which by default is set to `PhenoFlex_GDHwrapper`, a function included in the **chillR** package.

According to the help page for `chillR::phenologyFitter()`, the `modelfn` must be a *"[...] model function which computes the index in 'temperatures' at which blooming occures. It must have as first argument a data frame with at least the two columns 'Temp' and 'JDays' for one season, see 'SeasonList'. It can have further arguments which can be passed via '...'. The 'modelfn' must return a single numeric value for the predicted bloom JDay for that season. [...]"*

In short, the model function must calculate the bloom date for a data.frame `x` (with seasonal temperature data), given a specific parameter set `par`.  

Here's how the default wrapper function looks: 

```{r}
chillR::PhenoFlex_GDHwrapper
```

The function is relatively simple. It accepts: 

- `x`: a single-season data.frame (like those from `chillR::genSeasonList()`), with columns 'Temp', 'JDay' and 'Year'.
- `par`: a numeric vector of 12 parameters corresponding to the PhenoFlex model. 

It is important to maintain the correct order of parameters—if they are misaligned, function may produce incorrect results or fail altogether. 

The function performs three main tasks:

1. **Input checks:** It ensures that temperature parameters are valid (e.g. base temperature $T_b$ is lower than the optimal temperature $T_u$, and $T_u$ is less than the critical temperature $T_c$). 
2. **Model execution:** It runs the PhenoFlex model and retrieves the *bloomindex* (the row number where accumulated heat exceeds the requirement $z_c$). 
3.**Bloom date computation:** Converts the *bloomindex* into a Julian Day, factoring the hour of blooming to provide a more precise results. If the requirement is met at noon, the Julian Day is returned unchanged; otherwise, a fractional adjustment is applied.  

## Customizing Wrapper Function

The requirements for a `modelfn` wrapper a fairly loose —there’s no strict enforcement of how many parameters it must accept or in what order.

This flexibility allows us to **exclude certain parameter from calibration**, which can be useful when a parameter has limited influence on model performance or is already well understood. All we need to do is **manually fix its value inside the wrapper function**, and ensure it's excluded from `par.guess`, `lower` and `upper` when calling `chillR::phenologyFitter()`. 

Here's an example where we fix $T_c$ (critical temperature) at 36°C. Since it is no longer estimated, `par` will contain only 11 parameters:

```{r}
PhenoFlex_GDHwrapper_fixed <- function (x, par, Tc = 36) 
{
  par <- c(par[1:9], Tc, par[10:11])
    if (par[4] <= par[11]) 
        return(NA)
    if (par[10] <= par[4]) 
        return(NA)
    bloomindex <- PhenoFlex(temp = x$Temp, 
                            times = seq_along(x$Temp), 
                            yc = par[1], zc = par[2], 
                            s1 = par[3], Tu = par[4], 
                            E0 = par[5], E1 = par[6], 
                            A0 = par[7], A1 = par[8], 
                            Tf = par[9], Tc = par[10], 
                            Tb = par[11], slope = par[12], 
                            Imodel = 0L, 
                            basic_output = TRUE)$bloomindex
    if (bloomindex == 0) 
        return(NA)
    JDay <- x$JDay[bloomindex]
    JDaylist <- which(x$JDay == JDay)
    n <- length(JDaylist)
    if (n == 1) 
        return(JDay)
    return(JDay + 
             which(JDaylist == bloomindex)/n - 1/(n/ceiling(n/2)))
}

```


## Using the Custom Wrapper

Once your wrapper is defined, you can pass it to `chillR::phenologyFitter()` via `modelfn` argument. Be sure to:

- **Exclude** the fixed parameter $T_c$ from `par.guess`, `lower` and `upper`
- **Adjust the parameter indices** in the wrapper function
- **Use the modified wrapper** in the fitter function call

Here is an adapted example from the `chillR::phenologyFitter()` help page:

```{r, eval=FALSE}
data(KA_weather)
data(KA_bloom)
hourtemps <- stack_hourly_temps(KA_weather, 
                                latitude=50.4)
SeasonList <- genSeasonList(hourtemps$hourtemps, 
                            years=c(2007,2008))
#remove Tc from par.guess, upper and lower
#--> only 11 parameters to estimate
#order: yc, zc, s1, Tu, E0, E1, A0, A1, Tf, Tb, slope   
par <- c(40, 190, 0.5, 25, 3372.8, 9900.3, 6319.5, 
         5.939917e13, 4, 4, 1.6)
upper <- c(41, 200, 1, 30, 4000, 10000, 
           7000, 6.e13, 10, 10, 50)
lower <- c(38, 180, 0.1, 0, 3000, 9000, 
           6000, 5.e13, 0, 0, 0.05)
X <- phenologyFitter(par.guess=par, 
                     bloomJDays=KA_bloom$pheno[c(24,25)], 
  SeasonList=SeasonList, lower=lower, upper=upper,
  control=list(smooth=FALSE, verbose=TRUE, maxit=10, 
               nb.stop.improvement=5))
summary(X)
plot(X)

```


## Bonus: Override the Default Value via `...`

The best part about adding $T_c$ as an **optional argument** to your wrapper is that you can override its value directly when calling `chillR::phenologyFitter()`—without rewriting the wrapper.

```{r, eval=FALSE}
data(KA_weather)
data(KA_bloom)
hourtemps <- stack_hourly_temps(KA_weather, 
                                latitude=50.4)
SeasonList <- genSeasonList(hourtemps$hourtemps, 
                            years=c(2007,2008))
#remove Tc from par.guess, upper and lower
#--> only 11 parameters to estimate
#order: yc, zc, s1, Tu, E0, E1, A0, A1, Tf, Tb, slope  
par <- c(40, 190, 0.5, 25, 3372.8, 9900.3, 
         6319.5, 5.939917e13, 4, 4, 1.6)
upper <- c(41, 200, 1, 30, 4000, 10000, 
           7000, 6.e13, 10, 10, 50)
lower <- c(38, 180, 0.1, 0, 3000, 9000,
           6000, 5.e13, 0, 0, 0.05)
X <- phenologyFitter(par.guess=par, 
                     bloomJDays=KA_bloom$pheno[c(24,25)], 
  SeasonList=SeasonList, lower=lower, upper=upper,
  Tc = 30,
  control=list(smooth=FALSE, verbose=TRUE, maxit=10, 
               nb.stop.improvement=5))
summary(X)
plot(X)

```

This way, if you later want to fix $T_c$ at a different value, you don't need to edit the wrapper—just pass a new value via `$T_c=...$` in the function call. Much cleaner! 