---
title: "Implementing Alternative Calibration Algortihms"
output: distill::distill_article
bibliography: vignette_phenoflex.bib
csl: elsevier-harvard-modifed.csl
---

## Motivation

The chapters on fixing model parameters and introducing intermediate parameters showed that customizing the evaluation function is relatively simple. It gets a bit more complicated when you want to try out a different optimization algorithm than the simulated annealing function, implemented in the `phenologyFitter()` function. For example, I wanted to use the optimization framework `MEIGO` [@egea_meigo_2014]. `MEIGO` is also available on bioconductor, a platform where you can access R packages similar to CRAN. The installation via bioconductor is in my opinion less convenient than via CRAN and installing MEIGOR package was error-prone. As a quick fix, I mirrored the core of the MEIGOR package in my custom R package LarsChill (quite an original name, I know) <https://github.com/larscaspersen/addition_chillR>. Here you can find the original vignette for MEIGOR: <https://www.bioconductor.org/packages/release/bioc/manuals/MEIGOR/man/MEIGOR.pdf>. I only took the function for the Enhanced Scatter Search (ESS), which you can find via `LarsChill::custom_essr()`, but the functionality is just the same as in the original MEIGOR package. 

The ESS has some advantages and disadvantages compared to the simulated annealing algorithm. It was developed for complex optimization problems, it allows to screen for many different sets of model parameters, something I had the feeling the standard `phenologyFitter()` function was lacking. It is also relatively fast for small datasets (with <100 observations). You can specify additional inequality constraints when setting up the optimization problem. As you saw in the previous example, also the GDH_wrapper has some additional constraints implemented, but MEIGOR makes these easier to detect and to adjust and it also allows to specify acceptable ranges. I also like, that the function returns all the intermediate solutions that were surpassed by other sets of parameters in the course of the optimization run, it also returns the intermediate error scores, convenient if you want to make sure that the algorithm has converged or not. A big disadvantage is, that the speed of the ESS algorithm deteriorates quite fast for larger datasets, so if you have lots of observations in the calibration dataset it is better to stick with the simulated annealing algorithm used in the phenologyFitter function. 

## Implementing a new optimization algortihm for PhenoFlex is quite messy

To optimize PhenoFlex with another algorithm involves a little bit more work. The ESS optimization algorithm needs a function that calculates the model error for a particular set of parameters and temperature data. In most optimization problems model error is measured as Residual Sum of Squares (RSS), which is in our case simply the sum of the squared differences between predicted and observed bloom dates. The `PhenoFlex_GDH_wrapper()` function, however, only returns the bloom date for one season of temperatures and it does not even compare the prediction to the observed bloom dates. The function taking care of calculating the RSS is a hidden function called `chillR:::chifull()`. Note that I used three colons, when referring to the function, because this notation allows to access "hidden" functions, that did not get exported when creating the chillR package.  

I decided to maintain the modular structure of the `phenologyFitter()` function, where you specify the function how to calculate a bloom date with the `modelfn` argument, like with the `PhenoFlex_GDH_wrapper()` function. The evaluation function needs to do three things: 

- calculate bloom dates based on the temperature data specifeid in `SeasonList`
- calculate RSS by comparing the predicted bloom dates with the observed ones, specified in `bloomJDays`
- check model parameters for additional constraints

Here is an example of an evaluation function, that is compatible with the ESS algorithm of MEIGOR. I tried to 
bundle all the different cusotm evaluaiton functions and wrapper functions in a seperate package called
`evalpheno`, you can find the package here: <https://github.com/larscaspersen/eval_phenoflex>

```{r}
evalpheno::evaluation_function_meigo
```

As you can see, the evaluation functions is actually quite simple. At first we predict bloom dates, using the entries of `SeasonList`, t the function to calculate bloom dates specified in `modelfn` and the parameters specified in `x`(in the vignette for the MEIGOR package they always called the vector of the parameter `x`, while in PhenoFlex it is the name for the temperature data. Because I was not sure if the naming affected something, I followed the naming convention of the vignette, though it might be confusing.).

In case there is an NA in the predicted bloom dates, the value gets replaced with a value specified in `na_penalty`. Next, predicted and observed bloom dates get compared and the RSS is calculated and saved in the object `F`. The last and longest part is about the inequality constraints. The first three of the constrains should be familiar, as they are the same as in the PhemoFlex_GDH_wapper() function. Constrains 4 and 5 are new, relate to the $Q_{10}$ metric, that expresses by which factor a biological process speeds up for an increase in temperature by 10°C [@hegarty_temperature_1973]. In biological systems $Q_{10}$ values of 1.5 to 3.5 are deemed realisitic [@egea_reducing_2021]. In the early PhenoFlex papers [@luedeling_phenoflex_2021; @fernandez_unusually_2022] the $Q_{10}$ metric was checked *after* the calibration, but I think it makes even more sense to check the metric *during* the calibration. The results of the inequality constraints get saved in the vector `g`. Both, the error function `F` and the results of the inequality constrains `g` get returned to the optimization function. 

Before running the optimization function, we can specify acceptable upper and lower ranges for the inequality constrains, similar to how you do it for the search space of the model parameters. 

## Workflow using ESS algorithm

Here is an example of the specifications I made when calibrating PhenoFlex for a large dataset of temperate fruit trees [@caspersen_contrasting_2025]:

```{r, eval=FALSE}
#search space of model parameters
#        yc      zc     s1      Tu      theta_c   tau      piec    Tf     Tb     slope
x_0 <- c(24.79,	337.04,	0.2529,	17.72,	285.54,	   45.67,	  29.49,	2.97,	1.87,	2.69)
x_U <- c(80,    500,    1.0,    30,     287,       48,      50,    10,    10,     5.00)
x_L <- c(20,    100,    0.1,    15,     284,       16,      24,     2,     2,     1.2)

#limits for the inequality constraints
#         #gdh parameters   #q10 for E0 and E1
c_L <- c(  0,   0,   0,     1.5, 1.5)
c_U <- c(Inf, Inf, Inf,     3.5, 3.5)

problem<-list(f="evalpheno::eval_phenoflex_single_twofixed",
              x_0 = x_0,
              x_L = x_L,
              x_U = x_U,
              c_L = c_L, 
              c_U = c_U)


#options for fitter
opts<-list(#maxeval = 1000,
  maxtime = 60 * 10, 
  local_solver = 'DHC', 
  local_bestx = 1)

LarsChill::custom_essr(problem = problem,
                       opts = options,
                       modelfn = custom_PhenoFlex_GDHwrapper,
                       bloomJDays = pheno,
                       SeasonList = season_cal_list)
```

The first part is about the search space of the model parameters. I decided to optimize based on the intermediate model parameters and to fix two of the model parameters ($\theta^{*} = 279$, $T_c=36$). `x_0` is the initial guess, `x_L` the lower search range and `x_U` the upper range. The upper acceptable limit of the inequality constrain is specified in `c_U` and the lower limit in `c_L`. In the first three constrains for the heat submodel temperature parameters, we excluded any negative value, in the last two relating to the $Q_{10}$ metric, we excluded values outside the range of 1.5 to 3.5. All of that information is bundled in a list called `problem`. There it is also specified which **evaluation function** we are using, here I use the function `evalpheno::eval_phenoflex_single_twofixed`. Make sure that you omit the paranthesis after function name, because you want to only specify the name and not exexcute the function. The problem-list gets followed by the option list, here you specify settings of the optimization algorithm. `maxtime` specifies the time-limit for how long the algortihm should optimize (in seconds). `local_solver` specifies which optimization algorithm should be run when ESS found a set of model parameter leading to a lower RSS. The local solver makes a more refined narrower search around the newly found, better model parameters. `local_bestx` specifies how often the local search is called. When set to 1, local search is only triggered immediately after a better set is found, higher values would postpone the local search for some further iterations, that can be beneficial when the global search quickly finds even better parameter sets. You can also specifiy how many iterations the algorithms should make with `maxeval`. For more options, check the MEIGOR vignette. 

Lastly, the problem and option list get supplied to the `LarsChill::custom_essr()` funciton. This function does the optimization and should hopefully return good model parameters for your dataset. As you may have noticed, I also supplied three further arguments to the function: `modelfn`, `bloomJDays` and `SeasonList`. These are inputs for the evaluation function specified in the problem list. They are the same when running the `phenologyFitter()` function.

## The ouput


